\chapter*{Contributors}
\addcontentsline{toc}{chapter}{Contributors}
\markboth{Contributors}{Contributors}

% =====================================================
% Page 1 - Contributors Summary
% =====================================================

This page lists the main contributors of the Salam project.

\bigskip

\textbf{Salam GitHub Organization}

\url{https://github.com/SalamLang}

\bigskip

\textbf{Core Team}

\begin{description}[leftmargin=2.5cm]
\item[Max] Founder
\item[John Bampton] Lead Developer
\end{description}

\vfill

\begin{center}
\textit{"Salam is built by the community, for the community."}
\end{center}

\vspace{1cm}

\begin{center}
\small Artwork placeholder / Salam logo
\end{center}

\clearpage

% =====================================================
% Page 2 - Project Contributors
% =====================================================

\section*{Max}

\textbf{Founder}

\bigskip

\begin{itemize}
\item Placeholder biography sentence 1 about Max.
\item Placeholder biography sentence 2 about Max.
\item Placeholder biography sentence 3 about Max.
\item Placeholder biography sentence 4 about Max.
\item Placeholder biography sentence 5 about Max.
\end{itemize}

\bigskip

\begin{center}
Photo placeholder
\end{center}

\vfill
\hrule
\vspace{1cm}

\section*{John Bampton}

\textbf{Lead Developer}

\bigskip

\begin{itemize}
\item Placeholder biography sentence 1 about John.
\item Placeholder biography sentence 2 about John.
\item Placeholder biography sentence 3 about John.
\item Placeholder biography sentence 4 about John.
\item Placeholder biography sentence 5 about John.
\end{itemize}

\bigskip

\begin{center}
Photo placeholder
\end{center}

\vfill

\begin{center}
\small Salam Project --- Built with love and open collaboration.
\end{center}